The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2608 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f12521d29aa3827abb6a1d208e3d10c344e7607

Torsten Hothorn\\
Universit\"at Z\"urich \\
Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
Hirschengraben 84\\
CH-8001 Z\"urich, Switzerland \\
E-mail: \email{Torsten.Hothorn@R-project.org}\\
URL: \url{http://user.math.uzh.ch/hothorn/}\\

@@

A crucial difference between the testing strategies of CTree/MOB and GUIDE
is the difference between using only the residuals vs. the full model
scores. In the literature on structural change tests there is evidence
that residual-based tests can only capture parameter differences that
affect the conditional mean \citep[see e.g.,][ and further discussion in 
Section~\ref{sec:discussion}]{Ploberger+Kraemer:1992}.
Hence we compare the CTree, MOB, and GUIDE algorithms  -- all three using the 
default specification as shown in Table~\ref{tab:buildingblocks} -- and 
additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
dichotomized scores rather than residuals.

In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
represented by the corresponding selection probability, hence, a significance level 
is incorporated. To show that in general the drawn conclusions do not change if no 
significance level is included, a different illustration of the exact same evaluation 
is provided in Appendix~\ref{app:significance} where the performances are compared based 
on the smallest $p$-values, regardless of whether they are beneath any significance level.

Moreover, as defined by the true tree structure used in the data generating process, the
parameter functions $\beta_{k-1}(Z_1)$, $k=1,2=K$, are step functions which is particularly 
advantageous for the maximum selection in the test statistic applied by MOB while the linear 
selection in CTree is preferable for detecting monotonous dependencies. To show that employing 
a continuous parameter function instead of a step function only has an impact on whether CTree or
MOB performs better the exact same simulation setting as used for Figure~\ref{fig:scores_stump} 
but with continuously changing parameters has been evaluated and is illustrated in 
Appendix~\ref{app:contbeta}.

@@

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for the median $0$ and varying $\beta_0$), 
GUIDE is never ahead of them in any of the considered settings. Moreover, the positive 
effect of including all available scores can be observed particularly well in the top middle 
panel where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to GUIDE. 
%The fact that residual-based methods fail to detect changes that do not effect the
%expected value has already been elaborated in the literature of structural change
%analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
%Section~\ref{sec:discussion}. 
Furthermore, this advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
perform similarly well as long as the intercept varies (left and right panel) with CTree and 
MOB being only slightly ahead. However, for the split point at the $90$\% quantile $0.8$ 
(lower panels) the performance of all of the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a high effect 
size, MOB is clearly ahead with CTree leading to the second best results.
As discussed before, this advantage of MOB over CTree is mainly due to the abrupt shift in the 
model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
continuously changing parameters (see Appendix~\ref{app:contbeta}).
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and dichotomizing scores also decreases the
ability to detect the right split variable. To elaborate these effects more in detail 
the results for all possible combinations of the building blocks are discussed in a full 
factorial three-way analysis in the following section.

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the impact of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
considered. In the underlying evaluation all possible combinations of the building blocks 
have been included. The different levels of each of the three building blocks are listed in 
Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
In the case of categorization, the split variable $Z$ is binned at the
quartiles. This corresponds to a four-dimensional 0/1 transformation
function $g(Z)$ that indicates into which of the bins each observation falls.
Maximum selection across potential split points in a variable $Z$ also
corresponds to a multivariate 0/1 transformation function $g(Z)$. However,
in this case for each potential split point an indicator is used that is 0
before and 1 after the respective split point.
(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations of the building blocks.)

@@

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that dichotomizing 
residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
to the true split variable $Z_1$ and thus to lower power in the presented settings. 
For a central true split point ($\xi=0$, left panel) this effect remains at an almost constant 
level regardless of whether possible split variables are categorized or not. However, in case 
of the true split point at the $90$\% quantile $0.8$ (right panel), categorizing split 
variables increases the $p$-value of $Z_1$ even more. 
Here, the maximum selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes, but overall it depends on the situation whether a linear or a maximum
selection of a split variable is advantageous. 
Comparing the two panels suggests that unless one of the breaks used for binning 
possible split variables is close to the true split point (as for the median $0$, left panel), 
a categorization weakens the performance of a testing strategy. 
Moreover, while the effect of both of these transformations of the data (dichotomization of 
residuals/scores and categorization of split variables) can be observed
separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
remarkably.

@@

Note that the choice of the two specific settings 
(left panel: true split point at the median $0$,  effect size $0.3$; 
right panel: true split point at the $90$\% quantile $0.8$ requiring a higher effect 
size $1$)
has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
the differences particularly well observable. (Additional evaluations for increasing $\delta$ 
as in Section~\ref{sec:scores} can be found in Appendix~\ref{app:increasingdelta}.)

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       scales = list(alternating=FALSE),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                                   bg = "gray90"),
       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                  columns = 2,
                  lines = list(lty = c(0,1,2,0,1,1), 
                               type = "l", lwd = 2,
                               col = c("black","black","black","black", pal[1], pal[3]))))

@@

%\newpage
\subsection[``Tree'' scenario: Pre-pruning vs. post-pruning]{``Tree'' scenario: Pre-pruning vs.\ post-pruning}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$) both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.
The results presented in the top row of Figure~\ref{fig:tree_all} have been obtained applying pre-pruning.
Therefore, based on the corresponding ARI values conclusion can again be drawn on the power of the testing strategies 
as significance tests. CTree and MOB perform almost equally well for the true split points $0$, $0.2$ and $0.5$ 
($50$\%, $60$\%, and $75$\% quantile). GUIDE+scores performs slightly worse but it still achieves an ARI above 
$0.5$ for the highest considered effect size of $\delta = 1$. 
However, the original version of GUIDE is clearly behind which is mainly due to the fact that it doesn't detect 
the first split in $Z_2$ corresponding to a changing slope parameter as it only considers residuals in its testing 
strategy. 
For a split point farther away from the center such as the $90$\% quantile $0.8$ MOB 
outperforms all other methods and both GUIDE versions fail to find the correct tree structure.
Hence, the insights provided by the top row of Figure~\ref{fig:tree_all} conform with the conclusions
drawn from the ``stump'' scenario: Including all available scores in the testing strategy leads to a better performance 
while dichotomizing scores and categorizing split variables both do not cause any improvement but in many situations even 
have a negative effect on the power of a testing strategy.

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<tree_all, fig=TRUE, echo=FALSE, height=5, width=8>>=
library("latticeExtra")
useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
                      data = prex, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      key =  list(text = list(levels(prex$test)[1:4]),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(4,2),
                      scales=list(x=list(at=seq(1,11,2))),
                      xlab = expression(delta),
                      ylab = "Adjusted Rand Index",
                      #index.cond = list(c(2,3,1),c(2,1)),
                      par.settings = par.settings),
               strip = strip.custom(factor.levels = c(expression(xi==0), 
                                                      expression(xi==0.2), 
                                                      expression(xi==0.5), 
                                                      expression(xi==0.8)),
                                    bg = "gray90"),
               strip.left=strip.custom(factor.levels = c("Post-pruning", "Pre-pruning"),
                                   bg = "gray90"))
@
\caption{\label{fig:tree_all}
Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
``tree'' scenario, once applying pre-pruning (top row) and once post-pruning (bottom row). 
The ARI is evaluated over
increasing effect size $\delta$ with 100 replications of 250 observations per step.
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right, corresponding to the $50$\%, $60$\%, $75$\%, and $90$\% quantile).
}
\end{figure}

To elaborate the impact of the choice of a pruning strategy the presented tree models
are built again in the exact same simulation setting but employing post-pruning (i.e., building very 
large trees without any significance level and applying cost-complexity pruning as in the CART 
algorithm \citep{Breiman+Friedman+Stone:1984}). 
The corresponding ARI values are illustrated in the bottom row of Figure~\ref{fig:tree_all}
allowing for a comparison to those obtained with pre-pruning (top row).

For MOB and CTree it has been shown that their tests work very well as significance tests.
Therefore, it can be recommended to take advantage of this power and use the information
provided by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
in most cases the results of these two testing strategies do not vary significantly over 
a change of the applied pruning strategy.
On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
strategy should not be based on their results which argues for the application of post-pruning.
This is also supported by the results in Figure~\ref{fig:tree_all} showing that post-pruning improves
the performance of both GUIDE versions. Particularly if the true split point is close to a binning 
point (such that categorizing split variables has no negative effect) the post-pruning strategy 
compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely 
(bottom row, first and third panel).
In conclusion, the choice of a pruning strategy should be based on knowledge about the 
power of the applied testing strategy. With a well working significance test additional expensive
calculations such as needed for post-pruning may be avoided being one of the main advantages of 
pre-pruning. However, deficiencies of a testing strategy in terms of power can be at least partly
evened out by employing a post-pruning strategy.

>>>>

Torsten Hothorn\\
Universit\"at Z\"urich \\
Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
Hirschengraben 84\\
8001 Z\"urich, Switzerland \\
E-mail: \email{Torsten.Hothorn@R-project.org}\\
URL: \url{http://user.math.uzh.ch/hothorn/}\\

@@

A crucial difference between the testing strategies of CTree/MOB and GUIDE
is the difference between using only the residuals vs.\ the full model
scores. In the literature on structural change tests it is well-established
that residual-based tests can only capture parameter differences that
affect the conditional mean \citep[see e.g.,][ and further discussion in 
Section~\ref{sec:discussion}]{Ploberger+Kraemer:1992}.
Hence we compare the CTree, MOB, and GUIDE algorithms  -- all three using the 
default specification as shown in Table~\ref{tab:buildingblocks} -- and 
additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
dichotomized scores rather than residuals.

In Figure~\ref{fig:scores_stump} the performances of the testing strategies are 
represented by the corresponding selection probability, i.e., a significance level 
is incorporated. (Appendix~\ref{app:significance} shows that the same qualitative
conclusions can be drawn when the significance level is \emph{not} included.)
For a split point at the median ($\xi=0$, top row) all testing strategies 
perform similarly well as long as the intercept varies (left and right panel) with CTree and 
MOB being only slightly ahead. However, for the split point at the $90$\% quantile $\xi = 0.8$ 
(bottom row) the performance of all of the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a high effect 
size, MOB is clearly ahead with CTree leading to the second best results.
This advantage of MOB over CTree is mainly due to the abrupt shift in the 
model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
a smooth transition with continuously-changing parameters (see Appendix~\ref{app:contbeta}).

However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
It is easily possible, though, to substantially mitigate this problem by using scores
(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
is due to dichotomizing the scores at zero and due to categorizing the split variables
which are investigated in more detail in the following section.

@@

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}

To investigate the impact of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
considered. In this evaluation all possible combinations of the building blocks 
have been included. The different levels of each of the three building blocks are listed in 
Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}.

In the case of categorization, the split variable $Z$ is binned at the
quartiles. This corresponds to a four-dimensional 0/1 transformation
function $g(Z)$ that indicates into which of the bins each observation falls.
Maximum selection across potential split points in a variable $Z$ also
corresponds to a multivariate 0/1 transformation function $g(Z)$. However,
in this case for each potential split point an indicator is used that is 0
before and 1 after the respective split point.
(See also Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations of the building blocks.)

@@

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that dichotomizing 
residuals/scores decreases the performance as it leads to higher $p$-values for
the true split variable $Z_1$ and thus to lower power in the settings considered. 
For a true split at the median ($\xi=0$, left panel) this effect is almost constant
across the three types of categorization considered. However, in case 
of the true split point at the $90$\% quantile $\xi = 0.8$ (right panel), categorizing split 
variables increases the $p$-value of $Z_1$ even more. 
Not surprisingly, maximum selection is most advantageous in this case (i.e., for a late abrupt shift)
which is harder to detect based on a linear statistic. But overall it depends on the situation
whether a linear or a maximum selection of a split variable leads to lower $p$-values.

Comparing the two panels suggests that a categorization weakens the performance of the
tests unless the true split point is close to one of the breaks from the binning
(as in the left panel).
Moreover, while the effect of both transformations (dichotomization of residuals/scores and
categorization of split variables) can be observed separately in Figure~\ref{fig:3way},
combining them increases the negative impact remarkably.

@@

Note that the effect size $\delta$ in the results from Figure~\ref{fig:scores_stump} has been chosen
so that $p$-values for the non-dichotomized tests are roughly comparable: $\delta = 0.3$
for the true split point at the median $\xi = 0$ (left) vs.\ a stronger effect of $\delta = 1$
for the true split point at the 90\% quantile $\xi = 0.8$ (right).
Additional evaluations for varying effect size $\delta$ can be found in Appendix~\ref{app:increasingdelta}.

@@

\begin{figure}[t!]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       scales = list(alternating=FALSE),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                                   bg = "gray90"),
       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                  columns = 2,
                  lines = list(lty = c(0,1,2,0,1,1), 
                               type = "l", lwd = 2,
                               col = c("black","black","black","black", pal[1], pal[3]))))

@@

\subsection[``Tree'' scenario: Pre-pruning vs. post-pruning]{``Tree'' scenario: Pre-pruning vs.\ post-pruning}
\label{sec:tree}

So far the testing strategies underlying the different tree algorithms have only been considered
as classical significance tests, i.e., in terms of power and $p$-values. However, one could argue
that for a tree this is practically not really relevant -- at least when combined with a post-pruning
strategy such as cost-complexity pruning \citep{Breiman+Friedman+Stone:1984}. In the latter case it only matters that the relevant split
variables have the lowest $p$-value among all potential split variables -- but it is irrelevant whether
this is significant or not. To investigate to which extent this is actually true we evaluate the
different tree algorithms in the more complex ``tree'' scenario (see Figure~\ref{fig:dgp_tree}):
once with significance-based pre-pruning and once with cost-complexity post-pruning.

Recall that the true split structure is composed of splits in two different variables, both at
the same split point $\xi$. First, the split in $Z_2$ changes the slope from $+\delta$ to
$-\delta$. Second, the split in $Z_1$ changes the intercept in the negative slope group from
$-\delta$ to $+\delta$. In the simulation the effect size $\delta$ is increased from 0 to 1
for different split points $\xi$ from 0 to 0.8. Here, the performance is not evaluated in terms
of test properties but only in terms of tree properties, namely the adjusted Rand index (ARI)
in comparison to the true partition of the data.

The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
as the underlying significance tests have good power properties. Only for late abrupt shifts the
linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
both GUIDE flavors clearly perform worse as they underlying tests are less powerful; notably
for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<tree_all, fig=TRUE, echo=FALSE, height=5, width=8>>=
library("latticeExtra")
useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
                      data = prex, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      key =  list(text = list(levels(prex$test)[1:4]),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(4,2),
                      scales=list(x=list(at=seq(1,11,2))),
                      xlab = expression(delta),
                      ylab = "Adjusted Rand Index",
                      #index.cond = list(c(2,3,1),c(2,1)),
                      par.settings = par.settings),
               strip = strip.custom(factor.levels = c(expression(xi==0), 
                                                      expression(xi==0.2), 
                                                      expression(xi==0.5), 
                                                      expression(xi==0.8)),
                                    bg = "gray90"),
               strip.left=strip.custom(factor.levels = c("Post-pruning", "Pre-pruning"),
                                   bg = "gray90"))
@
\caption{\label{fig:tree_all}
Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
``tree'' scenario, once applying pre-pruning (top row) and once post-pruning (bottom row). 
The ARI is evaluated over
increasing effect size $\delta$ with 100 replications of 250 observations per step.
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right, corresponding to the $50$\%, $60$\%, $75$\%, and $90$\% quantile).
}
\end{figure}

When switching from pre-pruning to post-pruning (bottom row of Figure~\ref{fig:tree_all})
it is indeed shown that many of the problems stemming from the low power of the two GUIDE
flavors are indeed mitigated. Thus, by first growing a large tree without stopping upon
non-significance and then pruning back based on predictive performance instead substantially
improves the fit of the two GUIDE flavors. However, using the residuals only in GUIDE still
performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
CTree and MOB when the true split matches one of the categorization bins (i.e., $xi = 0$
and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
cases ($\xi = 0.2$ and $\xi = 0.8$).

In summary, there is clear support for the conventional wisdom that the power of the testing
strategy in an unbiased tree algorithm is not so important when combined with post-pruning.
However, there are limits to this. Consequently, when tests can be made more powerful --
e.g., by using scores instead of residuals -- then this improves the corresponding tree
algorithm. Finally, the simulation also supports using pre-pruning in an unbiased tree algorithm
when the underlying testing strategy that also works well as a classical significance test.
